Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2018-11-22 documentation/paper.tex
Addressed lots and lots of comments

As part of the PICO-60 experiment, some testing was conducted to determine whether machine learning (in the form of a multi-layer perceptron) could be a viable technique to consider for a discrimination function. The results, while not definitive, indicated that the general technique warranted further investigation.

@@

\subsubsection{Piezo-Derived}

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item L2 Regularization $\lambda \in \{0.0003, 0.001, 0.003\}$
    \item Dense Layer Dropout $\in \{0, 0.25, 0.5\}$
    \item Convolutional Filters $\in \{24, 48\}$
    \item Convolutional Kernel Size $\in \{3, 5\}$
    \item Number of Convolutional Layers $\in \{17, 32\}$
    \item Number of Dense Layers $= 3$
\end{itemize}

@@

During early experiments with shallower network architectures, it was revealed to have a surprising and undesirable capacity to separate run types based on meaningless information. It was able to discriminate between events from the AmBe neutron calibration runs and events from alpha background runs with 96\% accuracy, and to separate all combined neutron calibration runs from alpha background runs with 91\% accuracy. This was determined to be caused by the network detecting and fitting on biases in the white noise at the beginning and hydraulic recompression sounds at the end of the raw waveform. This extraneous information was subsequently cropped out.

When this extraneous information was cropped out and the architecture was modified, both accuracy and class-wise standard deviation remained quite high. The most accurate configuration during a grid search was 95\% accurate and had a standard deviation of 0.52 (seen in Figure \ref{waveform_hist}). This is better by both metrics than the multi-layer perceptron used in the original PICO-60 analysis.

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item Number of Layers $\in \{2, 3, 4\}$
\end{itemize}

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.006, 0.01\}$
    \item Number of Convolutional Layers $\in \{6, 9, 12\}$
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{image_hist}
    \caption{\label{image_hist} Prediction distribution of one of the best {\it ImageConv}$(\iota)$ discriminators.}
\end{figure}

@@

\begin{enumerate}
    \item This was a highly effective configuration, producing 97\% accuracy and a class-wise standard deviation dramatically better than the original PICO-60 neural network on that data set.
    \item The use of the exact same input format as AP makes the algorithms more directly comparable.
    \item If position correction is not applied, there is some risk of introducing a bias to the network's training data because of the different position distributions of the two classes.
\end{enumerate}

@@

Since the intent of these techniques is to separate particle types (the same as AP), performance evaluations (accuracy and class-wise standard deviation) use AP as a baseline. To ensure that accuracy statistics are as consistent and reliable as possible, every configuration was trained three times, with a different randomly selected validation set each time. (Using more than three training runs would have taken prohibitive amounts of time.) Accuracy, precision, recall, and class-wise standard deviation statistics are the mean over those three tests.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{grav_diff}
    \caption{\label{grav_diff} A visualization of $\mathrm{GravDiff}(p, \psi, g)$ with respect to $p$ and $\psi$.}
\end{figure}

@@

@@

Table \ref{semi_success_overview} directly compares performance of the most successful semi-supervised learning configurations, with AP provided as a baseline for comparison. Remember that ``Mean AP Accuracy'' refers to accuracy at replication of AP, rather than replication of the ground truths.

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Summary of the most successful semi-supervised learning configurations.} \label{semi_success_overview}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            Initial $\varsigma$ Size & Technique & Mean AP Accuracy & Mean Precision & Mean Recall & Mean CWSD \\
            \hline
            32 & GD & 83\% & 80\% & 99\% & 0.63 \\
            \hline
            32 & ICN & 95\% & 94\% & 99\% & 0.38 \\
            \hline
            64 & GD & 97\% & 96\% & 100\% & 0.18 \\
            \hline
            64 & ICN & 97\% & 96\% & 99\% & 0.28 \\
            \hline
            128 & GD & 99\% & 99\% & 99\% & 0.17 \\
            \hline
            128 & ICN & 98\% & 98\% & 100\% & 0.21 \\
            \hline
            256 & GD & 99\% & 99\% & 100\% & 0.10 \\
            \hline
            256 & ICN & 99\% & 99\% & 100\% & 0.15 \\
            \hline
            278 & PICO-60 ML & 80\% & 76\% & 100\% & 0.64 \\
            \hline
            $\infty$ & AP & 100\% & 100\% & 100\% & 0.41 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

@@

All programming for this study was done in Python 3. Keras \cite{keras}, running on a TensorFlow backend, was used for all machine learning tasks. NumPy \cite{numpy} and SciPy \cite{scipy} were used for linear algebra and signal processing. ROOT \cite{root}, scikit-image \cite{scikit-image}, and scikit-learn \cite{scikit-learn} were used for data loading and storage. Matplotlib \cite{matplotlib} was used for data visualization.

>>>>

As part of the PICO-60 experiment, some testing was conducted to determine whether machine learning (in the form of a multi-layer perceptron) could be a viable technique to consider for a discrimination function. The results, published in the 2017 PICO-60 paper \cite{pico}, indicated that the general technique warranted further investigation.

The neural network in this study used the same $\beta_{8}$ input format as was applied in this study (detailed in Section \ref{piezo_derived}), with pre-trigger noise subtracted. Position corrections (also in Section \ref{piezo_derived}) were not applied; rather, the position of the event in the vessel is additionally used as input. AP was used to restrict the data used as input to the neural network. In addition to neutrons and alpha particles, gamma events were used as training and testing data for this neural network.

Software for this experiment was implemented using the MATLAB Neural Network Toolbox \cite{matlab}.

@@

\subsubsection{Piezo-Derived} \label{piezo_derived}

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item L2 Regularization $\lambda \in \{0.0003, 0.001, 0.003\}$
    \item Dense Layer Dropout $\in \{0, 0.25, 0.5\}$
    \item Convolutional Filters $\in \{24, 48\}$
    \item Convolutional Kernel Size $\in \{3, 5\}$
    \item Number of Convolutional Layers $\in \{17, 32\}$
    \item Number of Dense Layers = 3 (with neuron counts 64, 16, and 1)
\end{itemize}

@@

The most accurate configuration during a waveform CNN grid search was 95\% accurate and had a standard deviation of 0.52 (seen in Figure \ref{waveform_hist}).

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item Number of Dense Layers $\in \{2, 3, 4\}$
\end{itemize}

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.006, 0.01\}$
    \item Number of Convolutional Layers $\in \{6, 9, 12\}$
    \item Number of Dense Layers = 3 (with neuron counts 64, 16, and 1)
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{image_hist}
    \caption{\label{image_hist} Prediction distribution of one of the best {\it ImageConv}$(\iota)$ discriminators. (Note the truncated X axis.)}
\end{figure}

@@

\begin{enumerate}
    \item This was a highly effective configuration, producing 96\% accuracy and a class-wise standard deviation dramatically better than the original PICO-60 neural network on that data set.
    \item The use of the exact same input format as AP makes the algorithms more directly comparable.
    \item If position correction is not applied, there is some risk of introducing a bias to the network's training data because of the different position distributions of the two classes.
\end{enumerate}

@@

Because all data in the PICO-60 experiment is labeled, unlabeled training data sets are randomly sampled from the 496 training examples remaining (after the validation set is selected).

Since the intent of these techniques is to separate particle types (the same as AP), performance evaluations (accuracy and class-wise standard deviation) use AP as a baseline. All graphs in this section use AP's predictions to define the sets of neutrons and alpha particles.

To ensure that accuracy statistics are as consistent and reliable as possible, every configuration was trained three times, with a different randomly selected validation set each time. (Using more than three training runs would have taken prohibitive amounts of time.) Accuracy, precision, recall, and class-wise standard deviation statistics are the mean over those three tests.

@@

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{grav_diff}
    \caption{\label{grav_diff} A visualization of $\mathrm{GravDiff}(p, \psi, g)$ with respect to $p$ and $\psi$.}
\end{figure}

@@

Table \ref{semi_success_overview} directly compares performance of the most successful semi-supervised learning configurations, with AP provided as a baseline for comparison. Remember that ``Mean AP Accuracy'' is calculated relative to AP's predictions, rather than replication of the ground truths.

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Summary of the most successful semi-supervised learning models.} \label{semi_success_overview}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            Initial $\varsigma$ Size & Technique & Mean AP Accuracy & Mean Precision & Mean Recall & Mean CWSD \\
            \hline
            32 & GD & 83\% & 80\% & 99\% & 0.63 \\
            \hline
            32 & ICN & 95\% & 94\% & 99\% & 0.38 \\
            \hline
            64 & GD & 97\% & 96\% & 100\% & 0.18 \\
            \hline
            64 & ICN & 97\% & 96\% & 99\% & 0.28 \\
            \hline
            128 & GD & 99\% & 99\% & 99\% & 0.17 \\
            \hline
            128 & ICN & 98\% & 98\% & 100\% & 0.21 \\
            \hline
            256 & GD & 99\% & 99\% & 100\% & 0.10 \\
            \hline
            256 & ICN & 99\% & 99\% & 100\% & 0.15 \\
            \hline
            278 & PICO-60 ML & 80\% & 76\% & 100\% & 0.64 \\
            \hline
            N/A & AP & 100\% & 100\% & 100\% & 0.41 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

@@

All programming for this study was done in Python 3. Keras \cite{keras}, running on a TensorFlow \cite{tensorflow} backend, was used for all machine learning tasks. NumPy \cite{numpy} and SciPy \cite{scipy} were used for linear algebra and signal processing. ROOT \cite{root}, scikit-image \cite{scikit-image}, and scikit-learn \cite{scikit-learn} were used for data loading and storage. Matplotlib \cite{matplotlib} was used for data visualization.
